Some examples in extended logic counting classes

### 2017-01-08 LICS/counting.tex
sc

In this section, we show that by syntactically restricting $\qso$ one can capture different function complexity classes. 
In other words, by using $\qso$ we can extend the theory of descriptive complexity~\cite{immerman2012descriptive} from decision problems to counting problems. 
For this, we first formalize the notion of \emph{capturing} a function complexity class to then show how to capture classes like $\shp$, $\fp$, and $\fpspace$.

Fix a relational signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a function complexity class over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a function complexity class. Then we say that {\em  $\FF$ captures $\CC$ over $\R$-structures} if the  following conditions are true:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Finally, we say that $\FF$ captures $\CC$ if $\FF$ captures $\CC$ over $\R$-structures for every relational signature $\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
Note that in Definition~\ref{def:cap} the function $f \in \CC$ and the formula $\alpha \in \FF$ must coincide in all the strings that encode $\R$-structures. Since we want to capture $\CC$ over a fix set of structures (e.g. graphs, matrices), it is natural to just consider strings that encodes $\R$-structures. 
This definition is fairly standard in descriptive complexity \cite{L04,immerman2012descriptive} and all notions for capturing complexity classes restrict $f \in \CC$ similarly.

Then, what functional complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$ who is one of the most well-known functional complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$ which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall that (see Section~\ref{sec:previous}) that every function class $\# \LL$ is contained in $\eqso(\LL)$. Thus, by~\cite{SalujaST95} it directly follows that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction is also direct and follows by showing that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{prop:capture-shP-LFP}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton et al.~\cite{ComptonG96} show that $\seso$ captures $\spp$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
Indeed, by using free SO-variables one will easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and by only using free FO-variables one cannot capture exponential functions definable in $\fp$ (e.g. $2^{n}$).
Therefore, it is impossible to capture classes like $\fp$ or $\fpspace$ if we follow previous approaches like~\cite{SalujaST95}.

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{I86,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

Another natural function complexity class is $\fpspace$~\cite{Ladner89}, the functional version of $\pspace$. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of an $\pspace$-machine. 
On the other hand, an $\fpspace$ machine can generate outputs of exponentially many bits. 
For this reason, the second-order sum is not enough for the quantitative layer of a logic for $\fpspace$, that is, one needs to consider second-order products to produce big numbers. 
This leads us to the fragment $\qso(\pfp)$ which naturally captures $\fpspace$. 
\begin{theorem} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}
The proof of the previous theorem follows the same lines than the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of an $\fpspace$ can be reconstructed by using $\pfp$ and second-order quantitative quantifiers.

From the proof of the previous theorem a natural question follows: what happen if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also proposes the class $\nfpspace$ of all functions computed by $\pspace$-machines whose output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{cor:capture-fpspace-poly}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}

The results in this section validate $\qso$ as the logical framework for extending the theory of Descriptive Complexity to function computational complexity. In the next sections, we study subfragments of $\eqso$ and, moreover, show how to go beyond $\eqso$ to capture other computational classes.

>>>>

In this section, we show that by syntactically restricting $\qso$ one can capture different counting complexity classes. 
In other words, by using $\qso$ we can extend the theory of descriptive complexity~\cite{immerman1999descriptive} from decision problems to computation problems. 
For this, we first formalize the notion of \emph{capturing} a complexity class of functions.
%, and then show how to capture classes like $\shp$, $\fp$, and $\fpspace$.

Fix a relational signature $\R = \{R_1, \ldots, R_k\}$ and assume that $\A$ is an ordered (finite) $\R$-structure with a domain $A = \{a_1, \ldots, a_n\}$.
%$ containing $n$ elements. 
Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$. 
%(that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). 
Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
%We define the class of all $\R$-functions, denoted by $\Func(\R)$, as the class of all functions $f: \ostr \rightarrow \bbN$.
%Given a function complexity class $\CC$ (i.e. $f: \Sigma^* \rightarrow \bbN$ for every $f \in \CC$), we say that a function $f \in \Func(\R)$ can be computed in $\CC$ if there exists $g \in \CC$ such that $f(\A) = g(\enc(\A))$ for every $\A \in \ostr$. 
%Note that the function $g$ outputs $f$ for encodings of structures and can behave arbitrarily otherwise.
We can now formalize the notion of capturing a counting complexity class.
%over $\R$-structures.
\begin{definition} \label{def:cap}
	Let $\FF$ be a fragment of $\qso$ and $\CC$ a counting complexity class. Then {\em  $\FF$ captures $\CC$ over the class of ordered $\R$-structures} if the  following conditions hold:
	\begin{enumerate}
		\item for every $\alpha \in \FF$, there exists $f \in \CC$ such that $\sem{\alpha}(\A) = f(\enc(\A))$ for every $\A \in \ostr[\R]$.

		\item for every $f \in \CC$, there exists $\alpha \in \FF$ such that   $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A \in \ostr[\R]$.
	\end{enumerate} 
	Moreover, {\em $\FF$ captures $\CC$ over ordered structures} if $\FF$ captures $\CC$ over the class of ordered $\R$-structures for every relational signature $\R$.
\end{definition}
%For the sake of simplification, we denote the first condition by $\FF \subseteq \CC$ and the second condition by $\CC \subseteq \FF$.
In Definition~\ref{def:cap}, function $f \in \CC$ and formula $\alpha \in \FF$ must coincide in all the strings that encode ordered $\R$-structures. Notice that this restriction is natural as we want to capture %Since we want to capture 
$\CC$ over a fix set of structures (e.g. graphs, matrices).
%, it is natural to just consider strings that encodes $\R$-structures. 
Moreover, this restriction is fairly standard in descriptive complexity \cite{immerman1999descriptive,L04}, and it has also been used in the previous work on capturing complexity classes of functions \cite{SalujaST95,ComptonG96}.
%all notions for capturing complexity classes restrict $f \in \CC$ similarly.

What counting complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$, a well-known and widely-studied counting complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$, which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall  that every function class $\# \LL$ is contained in $\eqso(\LL)$ (see Section~\ref{sec:previous}). Thus, it directly follows from~\cite{SalujaST95}  that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction of Proposition \ref{prop:capture-shP} is straightforward to prove, and it follows by the fact that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{prop:capture-shP-LFP}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton and Gr\"adel~\cite{ComptonG96} show that $\seso$ captures $\spp$, where $\eso$ is the existential fragment of $\so$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{prop:capture-spanP}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
Can we capture $\fp$ by using $\# \LL$ for some fragment $\LL$ of $\so$? A first attempt could be based on the use of a fragment $\LL$ of $\so$ that capture either $\ptime$ or $\nlog$~\cite{G92}. Such an approach fails as $\# \LL$ can encode $\shp$-complete problems in both cases; in the first case, one can encode the problem of counting the number of satisfying assignments of a Horn  propositional formula, while in the second case one can encode the problem of counting the number of satisfying assignments of a 2-CNF propositional formula. A second attempt could be based then on considering a fragment $\LL$ of $\fo$. 
But even if we consider the existential fragment $\Sigma_1$ of $\fo$ the approach fails, as $\# \Sigma_1$ can encode $\shp$-complete problems like counting the number of satisfying assignments of a 3-DNF propositional formula\cite{SalujaST95}. One last attempt could be based on disallowing the use of second-order free variables in $\sfo$. But in this case one 
%The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
%So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
%By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
%Indeed, 
%By using free second-order variables one can easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and 
%By only using free first-order variables one 
cannot capture exponential functions definable in $\fp$ such as~$2^n$.
%(e.g. $2^{n}$), so
Hence, it is not  clear how to capture
%impossible 
%to capture 
$\fp$ 
%or $\fpspace$ 
by following the approach proposed in~\cite{SalujaST95}. 
%if we follow previous approaches like
On the other hand, if we consider our framework and move out from $\eqso$, we have other options for counting like first- and second-order products. In fact, the combination of $\qfo$ with $\lfp$ is exactly what we need to capture $\fp$.
% the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To prove this theorem, 
%capture $\fp$ 
one first shows that every formula in $\qfo(\lfp)$ can be evaluated in polynomial time. 
Indeed, $\lfp$ is a polynomial-time logic~\cite{I86,vardi1982complexity}, and the sum and product quantifiers can also be computed in polynomial time. 
For the other direction, one has to use $\qfo(\lfp)$ to simulate the run of a polynomial time TM $M$ computing a function, in particular using the quantitative quantifiers to reconstruct the natural number returned by $M$ in the output tape. 
%built in the ouptape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for this reconstructions and, thus, crucial for capturing $\fp$.
%That is, without these operators is difficult to capture $\fp$.

At this point it is natural to ask whether one can extend the previous idea to capture $\fpspace$~\cite{Ladner89}, the class of functions computable in polynomial space. 
%Another natural function complexity class is $\fpspace$~\cite{Ladner89}, the functional version of $\pspace$. 
Of course, for capturing this class one needs a logical core powerful enough, like $\pfp$, for simulating the run of a polynomial-space TM.
% $\pspace$-machine. 
Moreover, 
%On the other hand, 
one also needs more powerful quantitative quantifiers as functions like $2^{2^n}$ can be computed in polynomial space,
%such a TM can generatean $\fpspace$ machine can generate outputs of exponentially many bits. 
so second-order sum is not enough for the quantitative layer of a logic for $\fpspace$.
%For this reason, the second-order sum is not enough for the quantitative layer of a logic for $\fpspace$, that is, one needs to consider second-order products to produce big numbers. 
In fact, by considering second-order product we obtain the fragment $\qso(\pfp)$ that 
%naturally 
captures $\fpspace$. 
\begin{theorem} \label{theo:capture-fpspace}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}
The proof of the previous theorem follows the same line as for the logical characterization of $\fp$: one shows that each function in $\qso(\pfp)$ can be computed in $\fpspace$ and, conversely, the output of a polynomial-space TM can be reconstructed by using $\pfp$ and second-order quantitative quantifiers.

From the proof of the previous theorem a natural question follows: what happens if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also introduced the class $\nfpspace$ of all functions computed by polynomial-space TMs 
%$\pspace$-machines 
with output length bounded by a polynomial.
% output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{cor:capture-fpspace-poly}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}

The results of this section validate $\qso$ as an appropriate logical framework for extending the theory of Descriptive Complexity to counting complexity classes. In the following sections, we provide more arguments for this claim, by considering some fragments of $\eqso$ and, moreover, by showing how to go beyond $\eqso$ to capture other 
%counting complexity 
classes.
